---
title: "Predictions using the Weight Lifting Exercises Dataset"
author: "Tamer Salem"
date: "5/4/2020"
output: html_document
---

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement ??? a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

## Data

The training data: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data source: http://groupware.les.inf.puc-rio.br/har.

# Initialization

```{r, echo=FALSE,}
library(caret)
library(rpart.plot)
set.seed(1234567)
```

Downloading and loadeding variable training_data and test_data respectively
```{r pressure, echo=TRUE, cache=TRUE}
training_data <- read.csv('pml-training.csv', na.strings = c("NA", "#DIV/0!", ""))
test_data <- read.csv('pml-testing.csv', na.strings = c("NA", "#DIV/0!", ""))
```

## Data Preparation

1st. remove records that contains more than 95% of the observation to be NA. We filter out those records. 
```{r}
clnColumnIndex <- colSums(is.na(training_data))/nrow(training_data) < 0.95
clean_training_data <- training_data[,clnColumnIndex]
```

We then verifying we have removed NA correctly
```{r}
colSums(is.na(clean_training_data))/nrow(clean_training_data)
colSums(is.na(clean_training_data))
```

We also remove col1 to col7 because they are not related to the model
```{r}
clean_training_data <- clean_training_data[,-c(1:7)]
clean_test_data <- test_data[,-c(1:7)]
```

We then partition the training data into training set and cross validation set
```{r}
inTrainIndex <- createDataPartition(clean_training_data$classe, p=0.75)[[1]]
training_training_data <- clean_training_data[inTrainIndex,]
training_crossval_data <- clean_training_data[-inTrainIndex,]
```

Now change the test data set into the same 
```{r}
allNames <- names(clean_training_data)
clean_test_data <- test_data[,allNames[1:52]]
```

# ML Algorithm - Decision Tree

```{r, cache=TRUE}
decisionTreeMod <- train(classe ~., method='rpart', data=training_training_data)
```

Predict with decision tree and output the confusion matrix. It seems like the result of the model is not ideal.
```{r}
decisionTreePrediction <- predict(decisionTreeMod, training_crossval_data)
confusionMatrix(training_crossval_data$classe, decisionTreePrediction)
```

plotting the decision tree
```{r}
rpart.plot(decisionTreeMod$finalModel)
```

# ML Algorithm - Random Forest
```{r, cache=TRUE}
rfMod <- train(classe ~., method='rf', data=training_training_data, ntree=128)
rfPrediction <- predict(rfMod, training_crossval_data)
confusionMatrix(training_crossval_data$classe, rfPrediction)
```

# Prediction
Now we use it to predict the test set
```{r}
predict(rfMod, clean_test_data)
```

# Conclusion

As we can we from the result:

- The random forest algorithem far outperforms the decision tree in terms of accuracy.
- We are getting 99.25% in sample accuracy, while the decision tree gives us only nearly 50% in sample accuracy